\chapter{Chromatic polynomial}

\section{Definitions and simple properties}

In the last chapter we discussed the $4$-coloring problem for planar graphs. In the 1930s Birkhoff and Whitney had the idea that, instead of constructing one $4$-colouring for a planar graph, one could instead find a formula for the \emph{number} of $4$-colourings, and then use algebraic, analytical or whatever other tools to somehow prove that the formula always returns a strictly positive answer. The plan didn't work; but it lead to the notion of the chromatic polynomial.

\begin{definition} For a graph $G$ we define the chromatic function $P_G(t)$, $P(G,t)$ or $P(t)$, as 
\begin{align*}
P_G(t)&=\#\text{ of vertex colourings of G with colours }\lbrace 1,\dots ,t\rbrace\\
&=\left|\left\lbrace c:V(G)\to \lbrace 1,\dots ,t\rbrace : c\text{ is a colouring}\right\rbrace\right|
\end{align*}
\end{definition}

Note that the definition does not require that all the colors actually appear in the coloring, i.e. $c$ need not be surjective.

\begin{example}
\begin{enumerate}
\item[$\circ$] If $G=\overline{K_n}$, $P_G(t)=t^n$.
\item[$\circ$] If $G=K_n$, $P_G(t)=t(t-1) \cdots (t-(n-1))=t^{\underline{n}}$. (FIGURE)
\item[$\circ$] If $G=P_n$, $P_G(t)=t(t-1)^{n-1}$.
\item[$\circ$] If $G=\emptyset$, $P_G(t)=1$.
\item[$\circ$] Consider cycles, for instance $G=C_5$. (FIGURE) The direct method ends up with a small complication: we didn't keep track of whether $1$ and $4$ are coloured differently. We will return to this example later.
\item[$\circ$] The chromatic number can now be redefined as $\chi(G)=\min \lbrace k\in\mathbb{N} : P_G(k)>0\rbrace$.
\item[$\circ$] The $4$-colour theorem is equivalent to the statement that $P_G(4)>0$ for a planar graph $G$.
\end{enumerate}
\end{example}

Let us now try to generalize some of those examples.

\begin{lemma}
If $G$ is a tree with $n$ vertices, then
\begin{align*}
P_G(t)=t(t-1)^{n-1}.
\end{align*}
\end{lemma}
\begin{proof}
If $G$ is a single vertex, then $P_G(t)=t$. Pick a leaf $x\in V(G)$. After $G\setminus x$ has been coloured we can pick the color for $x$ in $t-1$ ways, so
\begin{align*}
P_G(t)=P_{G-x}(t)(t-1) = t(t-1)^{n-2}(t-1)=t(t-1)^{n-1}
\end{align*}
by induction.
\end{proof}

\begin{lemma}
The chromatic polynomial of a disjoint union is $P(G\sqcup H,t)=P(G,t)P(H,t)$.
\end{lemma}
\begin{proof}
Any pair (colouring of $G$, colouring of $H$) gives a colouring of $G\sqcup H$.
\end{proof}

Next we show that the chromatic function is indeed a polynomial.
\begin{proposition}
\label{prop:chromatic-partition}
Let $\pi_i(G)$ be the number of ways to partition $V(G)$ into exactly $i$ non-empty independent sets. For any graph $G$ with $n>0$ vertices
\begin{align*}
P_G(t)=\sum_{i=0}^n \pi_i(G) t^{\underline{i}}.
\end{align*}
\end{proposition}
\begin{proof}
The choice of a colouring with colours from $\lbrace 1,\dots , t\rbrace$ is the same as 
\begin{enumerate}
\item[$\circ$] choosing the number $i$ of colours that will actually be used,
\item[$\circ$] partitioning $G$ into $i$ independent sets; there are $\pi_i(G)$ ways to do this,
\item[$\circ$] colouring each part with a different colour; there are $t(t-1) \cdots  (t-(i-1))=t^{\underline{i}}$ ways.
\end{enumerate}
\end{proof}
\begin{example}
Let $G=P_3$. (FIGURE)
\begin{align*}
\pi_1 &=0\\
\pi_2 &= 1\quad ,\qquad 1-2-1\\
\pi_3 &= 1\quad ,\qquad 1-2-3\\
\pi_{\ge 4} &= 0
\end{align*}
Then
\begin{align*}
P_{P_3}(t)=\pi_2t^{\underline{2}}+\pi_3 t^{\underline{3}}=t(t-1)+t(t-1)(t-2)=t(t-1)^2
\end{align*}
\end{example}

From now on we call $P_G(t)$ the \emph{chromatic polynomial}. There are other ways to express $P_G(t)$ which also certify that the function is polynomial.

\begin{proposition}
\label{prop:chromatic-inclusion}
Suppose $G=(V,E)$ is a nonempty graph. Then
\begin{align*}
P_G(t)=\sum_{F\subseteq E} (-1)^{|F|} t^{c(F)}
\end{align*}
where $c(F)$ is the number of connected components of $(V,F)$.
\end{proposition}

Before the proof we quickly recall the Inclusion-Exclusion principle.

\begin{fact}
Suppose $A$ is a set, and $A_1,\dots, A_k$ are subsets of $A$. For any $X\subseteq \lbrace 1,\dots , k\rbrace$ define
\begin{align*}
A_X=\bigcap_{i\in X} A_i
\end{align*}
Then
\begin{align*}
\left| \overline{\bigcup_{i=1}^k A_i}\right| = \sum_{X\subseteq \lbrace 1,\dots , k\rbrace} (-1)^{|X|}|A_X|
\end{align*}
\end{fact}

\begin{example} Assume we have some ambient set $\mathcal{A}$, with  $A,B,C \subset \mathcal{A}$. (FIGURE)
Then we have
\begin{align*}
|A\cup B \cup C| &= |A|+|B|+|C| -|A\cap B|-|B\cap C|-|C\cap A| + |A\cap B\cap C|\\\\
|\overline{A\cup B\cup C}|&= |\mathcal{A} | - |A|-|B|-|C| +|A\cap B|+|B\cap C|+|A\cap C| -|A\cap B\cap C|
\end{align*}
\end{example}

\begin{proof}[Proof of Proposition~\ref{prop:chromatic-inclusion}]
Define $A=\lbrace g:V\to \lbrace 1,\dots ,t\rbrace\rbrace$ (this is the set of all functions, not just colourings). For every $e=xy\in E$, let $A_e=\lbrace g\in A : g(x)=g(y)\rbrace$. Now for $F\subseteq E$, let $A_F= \bigcap_{e\in F} A_e$. Clearly $|A_F|=t^{c(F)}$, because $g\in A_F$ must be constant on every component of $(V,F)$. But then we're done, as 
\begin{align*}
P_G(t)= \left| \overline{\bigcup_{e\in E} A_e}\right| \overset{Incl.-excl.}{=} \sum_{F\subseteq E} (-1)^{|F|} |A_F|
\end{align*}
\end{proof}

This will help us understand some coefficients of the chromatic polynomial.

\begin{notation}
If $P(t)$ is a polynomial, we write $[t^k]P(t)$ for the coefficient of $t^k$ in $P(t)$.
\end{notation}
\begin{lemma}
If $G$ is a graph with $n$ vertices and $m$ edges, then $P_G(t)$ is a polynomial of degree $n$, with
\begin{align*}
[t^n]P_G(t)=1\quad , \qquad [t^{n-1}]P_G(t)=-m
\end{align*}
\end{lemma}
\begin{proof}
By Proposition~\ref{prop:chromatic-inclusion} we have $c(F)\le n \Rightarrow \deg P_G \le n$. Now then $c(F)=n $ if and only if $F=\emptyset$, which implies $t^n$ appears exactly once in $P_G(t)$ with coefficient $(-1)^{|\emptyset|}=1$. Also, $c(F)=n-1$ if and only if $F=\lbrace e\rbrace$ is a single edge, which implies $t^{n-1}$ appears with coefficient $(-1)^{-1}m=-m$.
\end{proof}

\section{Deletion--contraction}

\begin{notation}
For $e\in E(G)$ we write: $G-e$ for $G$ with $e$ removed, and $G/e$ for $G$ with $e$ contracted. \\
(FIGURE)
\end{notation}
\begin{proposition}
(Deletion--contraction rule). If $e\in E(G)$, then
\begin{align*}
P_G(t)=P_{G-e}(t)-P_{G/e}(t)
\end{align*}
\end{proposition}
\begin{proof}
Let $e=xy\in E(G)$.
\begin{align*}
P_{G-e}(t)&=\#\text{ colourings with } c(x)\neq c(y) + \# \text{ colourings with } c(x)=c(y) \\
&= P_G(t)+P_{G/e}(t)
\end{align*}
\end{proof}
\begin{remark}
$|E(G-e)|=|E(G)|-1$ and $|V(G/e)|=|V(G)|-1$, so we could define $P_G(t)$ recursively
\begin{align*}
P_G(t) = \left\lbrace \begin{array}{ll}
t^{|V(G)|} & \text{if } G \text{ has no edges}\\
 & \\
P_{G-e}(t)-P_{G/e}(t) & \text{if } e\in E(G)
\end{array} \right.
\end{align*}
\end{remark}

\begin{example} $G$ is a graph on $5$ vertices, and we run the above.
(FIGURE)
\begin{align*}
= t(t-1)^3-2t(t-1)^2(t-2)(t-2)+t(t-1)(t-2)
\end{align*}
\end{example}
\begin{example}
$P(C_n,t)=P(P_n,t)-P(C_{n-1},t)$. We have
\begin{align*}
P(P_n,t)&= t(t-1)^{n-1}\\
P(C_3,t)&=t(t-1)(t-2)
\end{align*}
Now one can prove by induction that $P(C_n)=(t-1)^n+(-1)^n(t-1)$. In particular
\begin{align*}
P(C_n,2)=1+(-1)^n= \left\lbrace \begin{array}{ll}
2, & 2\mid n\\
0, & 2\nmid n
\end{array}\right.
\end{align*}
which agrees with what we know about $\chi(C_n)$.
\end{example}

The deletion-contraction rule has many interesting consequences. 

\begin{proposition}
Let $G$ be a graph with $n>0$ vertices, $m$ edges and $c$ connected components. The coefficients of $P_G(t)$ alternate in signs, i.e.
\begin{align*}
P_G(t)=\sum_{i=0}^n (-1)^i c_i(G)t^{n-i}
\end{align*}
where $c_i(G)\ge 0$. Moreover $c_i(G)=0$ for $i>n-c$ and $c_{n-c}(G)\neq 0$.
\end{proposition}
Simply put, this proposition states that the chromatic polynomial has the form
\begin{align*}
P_G(t)=t^n-mt^{n-1}+c_2(G)t^{n-2}-\cdots + (-1)^{n-c}c_{n-c}(G)t^c.
\end{align*}
It is very easy to come up with a proof by induction, and we leave it as an exercise. It will be instructive for the future to phrase the proof slightly differently.
\begin{proof} Deleting an edge keeps the sign, and contracting an edge changes the sign. Successive contraction--deletion splittings produce a tree which looks like this:

(FIGURE)

We chose to continue with the splittings until we reach a graph with no edges, that is one of the $\overline{K_i}$, $1\le i \le n$. Every branch ending with $\overline{K_i}$ contributes
\begin{align*}
(-1)^{n-i}t^i
\end{align*}
to the chromatic polynomial. That is because the path from $G$ to $\overline{K_i}$ contains $n-i$ contractions, i.e. $n-i$ sign changes. It means that the proposition holds with
\begin{align*}
c_i(G)=\# \text{ of branches ending with } \overline{K_i}
\end{align*}
and clearly $c_i(G)\geq 0$.

To prove that $c_i(G)=0$ for $i>n-c$ note that no branch of the tree ends with a graph on less than $c$ vertices. Moreover, there is at least one branch with ends exactly with $K_c$ (apply contractions all the time), so $c_{n-c}(G)>0$.
\end{proof}


\section{Coefficient sequence}

As a small side remark, the behaviour of the coefficient sequences of chromatic polynomials is the subject of active research and is the topic of many conjectures and theorems. Most of that work takes place in the general context of \emph{matroids} and their \emph{characteristic polynomials}, of which the chromatic polynomial is a special case. Here is a very fresh result which proves a conjecture of Read from 1968. The proof uses techniques from algebraic geometry and singularity theory.

\begin{theorem} \emph{(June Huh, 2010)}
Suppose $G$ is connected with chromatic polynomial
$$P_G(t)=t^n-c_1t^{n-1}+c_2t^{n-2}-\cdots +(-1)^{n-1}c_{n-1}t.$$
Then the sequence $(1,c_1,c_2,\dots,c_{n-1})$ is log-concave, which means
$$c_{i-1}c_{i+1}\leqslant c_i^2 \quad \text{for all } i.$$
In particular, it is unimodal, which means
$$1 \leqslant c_1 \leqslant c_2 \leqslant \cdots \leqslant c_{k-1} \leqslant c_k \geqslant c_{k+1} \geqslant \cdots \geqslant c_{n-1}, \quad \text{for some } k.$$
\end{theorem}


\begin{remark} We can prove $1 \leqslant c_1 \leqslant c_2 \leqslant \cdots \leqslant c_{\lfloor \frac{1}{2}(n-1)\rfloor}$.
\\ If $G$ is a tree, then
$$P_G(t)=t(t-1)^{n-1}=\sum_{i=0}^{n-1} {n-1 \choose i}(-1)^it^{n-i}\cdot t=t^n-{n-1 \choose 1}t^{n-1}+{n-1 \choose 2}t^{n-2}-\cdots.$$
The sequence $(1,c_1,c_2,\dots)$ is $(1,{n-1 \choose 1},{n-1 \choose 2},\cdots)$, and it is increasing up to the middle term.
\\ Now suppose that $G$ is connected, but not a tree. Then, by definition of a tree, there is an edge $e\in E(G)$ such that $G-e$ is still connected. For $i\leqslant \frac{1}{2}(n-1)$ we notice that
$$P_G(t)=P_{G-e}(t)-P_{G/e}(t) \Longrightarrow c_{i-1}(G)=c_{i-1}(G-e)-(-c_{i-2}(G/e))=c_{i-1}(G-e)+c_{i-2}(G/e).$$
We know $i\leqslant \frac{1}{2}(n-1)$ and $i-1 \leqslant \frac{1}{2}(n-2)=\frac{1}{2}(|V(G/e)|-1)$, hence by induction 
 $$c_{i-1}(G)\leqslant c_i(G-e)+c_{i-1}(G/e)=c_i(G)$$
which ends the induction step.
\end{remark}


\section{Counting acyclic orientations}

What else does the chromatic polynomial count? And how?

\begin{definition} An \emph{orientation} of $G$ is a choice  of direction for every edge. This gives a directed graph. If $G$ has $m$ edges, then it has $2^m$ possible orientations (which might also be isomorphic).
\end{definition}

\begin{definition} An orientation is \emph{acyclic} if it has no closed directed walk. Let $a(G)$ be the number of acyclic orientations of $G$.
\end{definition}

\begin{theorem} \emph{(Stanley, 1973)}
If $G$ has $n$ vertices, then $a(G)=(-1)^nP_G(-1)$.
\end{theorem}

\begin{example}
\begin{itemize}
\item $G$ is a tree with $n$ vertices, then
$$a(G)=2^{n-1}=(-1)^n(-1)(-1-1)^{n-1}=(-1)^nP_G(-1),$$
\item $G$ is a cycle on $n$ vertices, then
\begin{align*}
a(G)&=2^n-2, \\
(-1)^nP_G(-1)&=(-1)^n[(-2)^n+(-1)^n(-2)]=(-1)^n[(-1)^n(2^n-2)]=a(G),
\end{align*}
\item $G=K_n$, then
\begin{align*}
(-1)^nP_G(-1)&=(-1)^n(-1)^{\underline{n}}=(-1)^n(-1)(-1-1)(-1-2)\cdots (-1-(n-1))=(-1)^n(-1)^nn!.
\end{align*}
An acyclic orientation is the same as ordering the vertices $v_1,v_2,\dots,v_n$ (there are $n!$ possibilities to do this) and then choosing the orientation
$$v_i\longrightarrow v_j, \quad \text{whenever } i>j.$$
\end{itemize}
\end{example}

\begin{proof}
Take $e=xy \in E(G)$. Write $a^+(G-e),a^-(G-e),a^0(G-e)$ for the number of acyclic orientations of $G-e$ such that:
\begin{itemize}
\item There is a directed walk in $G-e$ from $x$ to $y$ ($a^+$),
\item There is a directed walk in $G-e$ from $y$ to $x$ ($a^-$),
\item There is no directed walk either way ($a^0$).
\end{itemize}

We prove a few claims about these quantities:
\begin{itemize} 
\item $a(G-e)=a^+(G-e)+a^-(G-e)+a^0(G-e)$. 
An acyclic orientation in $G-e$ cannot have directed walks $x\longrightarrow y$ and $y\longrightarrow x$ at the same time. These three sets are therefore disjoint and they give all the possibilities.

\item $a(G/e)=a^0(G-e)$. 
Take an orientation of $G-e$ with no walk $x\longrightarrow y$ or $y\longrightarrow x$. For any $z\in N_{G-e}(x)\cap N_{G-e}(y)$, the edges $xz$ and $yz$ have the same orientation (if not, there would be a walk $x\longrightarrow z\longrightarrow y$ or $y\longrightarrow z\longrightarrow x$), hence either  
$$x\longrightarrow z \text{ and } y\longrightarrow z$$
or
$$z\longrightarrow x \text{ and } z\longrightarrow y.$$
The orientation of $G-e$ determines then an orientation of $G/e$ (the edges $xz$ and $yz$ are compatible under the contraction). This orientation is also acyclic (a directed walk from $xy$ to itself would imply a directed walk in $G-e$ from $x$ or $y$ to $y$ or $x$). This also works vice versa.
\\ The idea here was that 
$$\text{Closed walks in } G/e = \text{Walks } x\longrightarrow y \text{ or } y\longrightarrow x \text { in } G-e.$$

\item $a(G)=a^+(G-e)+a^-(G-e)+2a^0(G-e)$. 
For the first two terms there is only one way to extend the orientation of $G-e$ without closing a cycle in $G$. In the last case the edge $xy$ can be oriented both ways, since we don't have a walk from $x$ to $y$ or from $y$ to $x$.
\end{itemize}

By these three claims we obtain
\begin{align*}
a(G)&=a^+(G-e)+a^-(G-e)+2a^0(G-e)= \\
&= a^+(G-e)+a^-(G-e)+a^0(G-e)+a^0(G-e)= \\
&= a(G-e)+a^0(G-e)= \\
&= a(G-e)+a(G/e)
\end{align*}
We complete the proof by using induction:
\begin{itemize}
\item $G=K_1$, then $a(G)=1=(-1)^1P_{K_1}(-1)$,
\item Pick an edge $e\in E(G)$, then (by induction assumption)
\begin{align*}
a(G)&=a(G-e)+a(G/e)= \\
&= (-1)^nP_{G-e}(-1)+(-1)^{n-1}P_{G/e}(-1)= \\
&= (-1)^n[P_{G-e}(-1)-P_{G/e}(-1)]= \\
&= (-1)^nP_G(-1)
\end{align*}
\end{itemize}
\end{proof}

Stanley's theorem is, again, just the tip of an iceberg known as \emph{duality} in matroid theory. It has many generalizations.


\section{Chromatic roots}

Let's try to wrap up this chapter by going back to the original question of Birkhoff and Whitney: is there something interesting to say about the set of points where the chromatic polynomial is positive? Or, alternatively, is there something interesting to say about their roots?

\begin{definition} $\alpha \in \mathbb{C}$ is a \emph{chromatic root} if $P_G(\alpha)=0$ for some graph $G$.
\end{definition}

Here are some very simple observations about chromatic roots.

\begin{observation}
\begin{enumerate}
\item Every natural number is a chromatic root,
\item For any $G$ different from the empty graph, $P_G(0)=0$,
\item For any $G$ with at least one edge, $P_G(1)=0$,
\item If $\alpha$ is a chromatic root, then so is $\alpha+1$ (because $P_{G+K_1}(\alpha+1)=(\alpha+1)P_G(\alpha)$),
\item The set of chromatic roots is countable (as a subset of the algebraic numbers).
\end{enumerate}
\end{observation}

The next proposition is less obvious (at least one of its parts). The proof is an appropriate adaptation of the deletion-contraction principle (again).

\begin{proposition} There is no chromatic root in $(- \infty, 0)\cup (0,1)$.
\end{proposition}

\begin{proof}
$\alpha <0$ is not a root of $P_G(t)$, since the coefficients of the polynomial have alternating signs.

Take $\alpha \in (0,1)$. Because $P_{G\sqcup H}(t)=P_G(t)P_H(t)$, it suffices to prove that $P_G(\alpha)\neq 0$ for any connected graph. Apply the deletion-contraction rule to $G$, in such a way that all the intermediate graphs are connected. At each step, either $G$ is a tree (and we stop splitting) or there is an edge $e\in E(G)$ such that $G-e$ is still connected.

A branch of this splitting process with $i$ contractions
\begin{itemize}
\item ends with an $(n-i)$-vertex tree,
\item introduces a sign of $(-1)^i$,
\item contributes $t(t-1)^{n-i-1}$ to $P_G(t)$.
\end{itemize}
Define $d_i$ as the number of branches ending with an $(n-i)$-vertex tree, then
$$P_G(t)=\sum d_i(-1)^it(t-1)^{n-i-1},$$
and of course we have $d_i\geqslant 0$.
Evaluate $P_G(\alpha)$ for $\alpha \in (0,1)$:
$$sgn\{d_i(-1)^i\alpha (\alpha-1)^{n-i-1} \}=(-1)^i\cdot 1\cdot (-1)^{n-i-1}=(-1)^{n-1},$$
which means that all monomials in $P_G(\alpha)$  evaluate to positive or all evaluate to negative, hence $P_G(\alpha) \neq 0$ as $d_i>0$ for at least one $i$. 
\end{proof}

This time we applied the deletion-contraction principle until we reached trees, and not edge-less graphs.

It actually gets much better.

\begin{theorem} \emph{(Jackson, Thomassen)}
There are no chromatic roots in $(- \infty, 0)\cup (0,1)\cup (1,\frac{32}{27}).$ Moreover, the constant $\frac{32}{27}$ is optimal. 
\end{theorem}

It is also known (Sokal) that chromatic roots are a dense subset of the complex numbers. In the context of planar graphs the best we have now is:

\begin{theorem} \emph{(Birkhoff, Lewis)}
If $G$ is planar, then $P_G(t)>0$ for all $t\in [5,\infty)$.
\end{theorem}

It is conjectured that if $G$ is planar, then $P_G(t)>0$ for all $t\in [4,\infty)$. Even with the $4$-color theorem ($P_G(4)>0$) we still don't know if there is a \emph{planar} chromatic root in $(4,5)$. So, in a sense, Birkhoff's and Whitney's analytic--algebraic plan turned out to be much harder than the original combinatorial problem.

\section{Exercises}

\begin{enumerate}
\item What is $\min\{P_G(4)~:~G\ \mathrm{is\ planar}\}$\ ?
\item Find the chromatic polynomial of
\begin{itemize}
\item The $2\times n$ grid $K_2\square P_n$.
\item The wheel $W_n$.
\item $K_n-e$, the complete graph with one edge removed.
\item $P_n[K_m]$ (graph substitution, see TODO).
\item The disjoint union of two graphs.
\item $K_2 \square P_n$.
\end{itemize}
\item  Let $G,G_1,G_2$ be graphs such that $G=G_1 \cup G_2$ and $G_1 \cap G_2 \simeq K_k$ for some $k\geqslant 0$. Then
$$P_G(t)=\frac{1}{t^{\underline{k}}}P_{G_1}(t)P_{G_2}(t).$$
\item Let $P'_G(t)$ denote the number of colorings of $G$ with \emph{exactly} $t$ colors $1,\ldots,t$ appearing, i.e. such that the color assignment $c:V(G)\to\{1,\ldots,t\}$ is surjective. Show that $P'_G(t)$ is not a polynomial for any graph $G$.
\item If $G$ is a graph with chromatic polynomial $P(t)$, construct a graph with chromatic polynomial
\begin{itemize}
\item $tP(t-1)$,
\item $(t-1)P(t)$,
\item $P(t)^3$,
\item $P(t)^2/t$.
\end{itemize}
\item Let $G$ be a graph with $n$ vertices, $m$ edges and $r$ triangles. Show that $[t^{n-2}]P_G(t)={m\choose 2}-r$.
\item Let $G$ be a graph with $n$ vertices, $m$ edges and chromatic polynomial $P_G(t)=\sum_{i=0}^na_it^{n-i}$. Prove that $|a_i|\leq{m\choose i}$.
\item Where does the term \textit{log-concave} sequence come from? Prove that a log-concave sequence of positive real numbers is unimodal.
\end{enumerate}
