\chapter{Exam problems}

\section{Problems}

\begin{enumerate}

\item Find the chromatic number of the graph $G$ defined below in Sage:
\begin{verbatim}
def FunnyGraph(n):
    c = graphs.CompleteGraph(n)
    c.delete_edges(graphs.CycleGraph(n).edges())
    return graphs.MycielskiStep(c).join(graphs.WheelGraph(n+1))

G = FunnyGraph(99)
\end{verbatim}


\item Consider the following algorithm for vertex coloring. Find the largest independent set of vertices, and color them with color $1$. Remove those vertices, find the largest independent set in the remaining graph and color it with color $2$, and so on until there are no more vertices left to color. Prove that there are infinitely many graphs $G$ for which this algorithm will use more than $\chi(G)$ colors.


\item Prove that $\max\{\chi(G),\chi(\overline{G})\}\geq \sqrt{|V(G)|}$ for any graph $G$.


\item Let $G, H$ be two graphs. The \emph{substitution of $H$ into $G$}, denoted $G[H]$, is the graph obtained by replacing every vertex of $G$ with a copy of $H$, and replacing every original edge of $G$ with a complete bipartite graph between the corresponding copies of $H$. Formally $V(G[H])=V(G)\times V(H)$ and $(u,v)(u',v')\in E(G[H])$ iff either $uu'\in E(G)$ or $u=u'$ and $vv'\in E(H)$. Sage calls this operation \texttt{G.lexicographic\_product(H)}. Prove that
$$\omega(G)\chi(H)\leq \chi(G[H])\leq \chi(G)\chi(H).$$
Find an example with $\chi(G[H])<\chi(G)\chi(H)$.


\item Let $\textrm{gcd}(a,b)$ denote the greatest common divisor of $a$ and $b$. Let $n=20162016$. Define $G$ as the graph with vertex set $\{1,\ldots,n\}$ where two numbers $1\leq a<b\leq n$ are adjacent if and only if $\textrm{gcd}(a,b)=1$. Find the exact value of $\chi(G)$.


\item
\begin{itemize}
\item[a)] Prove or disprove: if the only complex roots of $P_G(t)$ are $0$ and $1$ then $G$ is a forest.
\item[b)] How many non-isomorphic graphs have chromatic polynomial $t^2(t-1)^8$\ ?
\item[c)] Find all non-isomorphic graphs with chromatic polynomial $t(t-1)^3(t-2)$.
\end{itemize}


\item Prove that $\chi_l(G)+\chi_l(\overline{G})\leq |V(G)|+1$ for any graph $G$, where $\chi_l$ is the list chromatic number.


\item 
(This is an experimental problem; formal proofs are not expected.) Let $g(n)$ be the expected number of colors used by the greedy algorithm to color a random graph from $G(n,\frac12)$. 
\begin{itemize}
\item Compute and plot an experimental approximation of $g(n)$ for a sequence of reasonably large values of $n$, for example $n=100,200,\ldots,2000$.
\item Speculate about the asymptotic behaviour of $g(n)$ as $n\to\infty$. In particular, what do you think about $\lim_{n\to\infty}\frac{g(n)}{n/\log_2{n}}$\ ?
\item Find information about the expected value of $\chi(G)$ for $G\in G(n,\frac12)$. How well does the greedy algorithm perform?
\end{itemize}


\item Let $G$ be a nonempty graph. Simplify the expression
$$\sum_I P(G-I, -1)$$
where the sum runs over all independent sets $I$ in $G$ (including the empty one) and, as always, $G-X$ denotes the subgraph of $G$ induced by the vertex set $V(G)- X$.


\item A vertex $v$ of a directed graph is called a \emph{source} if all the edges incident to $v$ are pointing out of $v$. Suppose $G$ is a nonempty graph with $n$ vertices. Prove that the number of acyclic orientations of $G$ having exactly one source equals $n\cdot (-1)^{n-1}\cdot [t]P_G(t)$.


\item Find the edge-chromatic number $\chi'$ of \texttt{FunnyGraph(99)} from Problem 1.


\item In the lectures we defined a family of graphs $Q_d(u,s)$, and we used the fact that they are unit distance graphs in $\real^d$. 
\begin{itemize}
\item[a)] Show that each $Q_d(u,s)$ is in fact a unit distance graph in $\real^{d-1}$. 
\item[b)] Use the graphs $Q_{10}(u,s)$ to prove $\chi(\real^9)\geq C$ for a constant $C$ as large as you can. 
\end{itemize}


\item The \emph{supremum metric} (or $\ell_\infty$ metric) in $\real^d$, $d\geq 1$ is given by
$$d_\infty((x_1,\ldots,x_d),(y_1,\ldots,y_d))=\max\{|x_1-y_1|,\ldots,|x_d-y_d|\}.$$
Find the smallest number of colors required to color $\real^d$ so that any two points whose distance in the supremum metric equals $1$ have different colors.


\item Let $P_{2\times n}=P_2\square P_n$ be the $2\times n$ grid graph. Find the number of edge-colorings of $P_{2\times n}$ with $3$ colors. ($P_{2\times n}$ is called \texttt{graphs.Grid2dGraph(2,n)} in Sage).
\end{enumerate}


\section{Hints}

\begin{enumerate}

\item The graph is $M(K_n\setminus C_n) + C_n + K_1$ and so its chromatic number is $(1+\lceil\frac{n}{2}\rceil) + (2+(n \mod 2)) + 1$ for $n>3$.

\item Take two $n$-vertex stars and connect their central vertices with an edge.

\item $\chi(G)\chi(\compl{G})\geq \chi(G)\alpha(G)\geq |V(G)|$.

\item For the lower bound $G$ contains an $\omega(G)$-fold join of copies of $H$. For the upper bound take a cartesian product of colorings of $G$ an $H$. For the example take $G=C_5$, $H=K_2$.

\item Let $\pi(n)$ be the number of prime numbers up to $n$. The graph contains a clique of size $1+\pi(n)$. There is also a coloring with $1+\pi(n)$ colors: map each number to the smallest prime in its factorization.

\item 

\item

\item

\item

\item

\item Find the degree sequence. There is only one vertex of maximal degree.

\item All vertices lie in the same $(d-1)$-hyperplane in $\real^d$. We have $\alpha(Q_{10}(4,5))=12$, so $\chi(\real^9)\geq \chi(Q_{10}(4,5))\geq {10 \choose 5}/12 = 21$. This proof comes from \url{http://arxiv.org/abs/1409.1278}.

\item A coloring $c:\real^d\to\{0,1\}^d$ given by
$$c(x_1,\ldots,x_d)=(\lfloor x_1\rfloor \ \textrm{mod}\ 2,\ldots, \lfloor x_d\rfloor \ \textrm{mod}\ 2 )$$
uses $2^d$ colors and is optimal.

\item There are many ways to observe inductively that $c_n=2c_{n-1}-6$ for $n\geq 3$, where $c_n=P(L(P_{2\times n}), 3)$.


\end{enumerate}
