%
\documentclass[a4paper]{article}


\usepackage{color}              %Farben, f.r \definecolor{}
\usepackage{amssymb}            %Mathematische Symbole
\usepackage{amsthm}             %Besseres \newtheorem
\usepackage{amsmath}           %Mathematische Umgebungen
\usepackage{mathtools}          %\xRightarrow, etc
\usepackage{mathrsfs}           %enthaelt \mathscr
\usepackage[matrix,arrow,curve]{xy}     %Diagramme
\usepackage{graphicx}
\usepackage{enumerate}          % in-place numerations def.
\usepackage{fullpage}

%% Dokument Beginn %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\pagestyle{empty}
\begin{center}
	{\Large\bf Graph coloring}\\
	{\large\bf Exercise class problems - volume 1}\\
	\line(1,0){330}
\end{center}

\begin{enumerate}
\item Prove that every graph has two vertices of the same degree.
\item Six people meet at a party. Prove that (1) there are three people that know each other or (2) there are three people that pairwise don't know each other.
\item In a class with $19$ students each person sends a Valentine's Day card to exactly three other students. Is it possible that each student receives cards from the same three students to whom he/she sent cards?
\item Draw all isomorphism classes of trees on $n=1,2,3,4,5$ vertices.
\item How many non-isomorphic trees on 10 vertices are there? (The answer is available online)
\item What is the maximal possible number of edges in a disconnected graph on $n$ vertices?
\item Let $G$ be a graph with minimum vertex degree $d\geq 2$. Prove that $G$ contains a cycle of length at least $d+1$.
\item Prove that if $G$ is disconnected then its complement $\overline{G}$ is connected.
\item Classify all connected $2$-regular graphs.
\item Prove that the $k$-cube $Q_k$ is regular (of what degree)?
\item Find $\alpha(Q_k)$.
\item For which $n$ is there a homomorphism $C_n\to K_2$\ ?
\item For which $n,m$ is there a homomorphism $C_n\to C_m$\ ?
\item Construct a graph $G$ such that the only homomorphism $G\to G$ is the identity.

\end{enumerate}


\end{document}




